---
title: "Ideas_2"
author: "Michael Lockhart"
date: "11/5/2019"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**IDEAS**
*We can compare ticket sale, merch sale, attendance, players, player pay, aggressive-ness, etc for womens and mens soccer.*
*We can explore one sport fully and look at how many people watch the game live in the stadium, which game had the highest attendance,Who is the most experienced player in the game, What’s the winning rate for each team, What’s the relationship between aggressive behaviors and the winning rate.*
```{r}
library(readr)
library(tidyverse)
library(lubridate)
library(ggmap)
library(babynames)
library(plotly)
library(gganimate) 
library(gifski) 
```

```{r}
nfl_elo <- read_csv("https://projects.fivethirtyeight.com/nfl-api/nfl_elo.csv")
```

```{r}
nfl_elo %>% 
  filter(team1 == "MIN") %>% 
  arrange(desc(elo1_pre)) %>% 
  
  ggplot(aes(y = elo1_pre, x = season))+
  geom_point() +
  geom_smooth()
```

Questions for Research:

What does "elo1_pre" mean? We assume it is a metric five thirty eight loses for their NFL predictions

What caused the peaks and valleys in the trends? Clearly at their conception, the team started off slowly in terms of the elo1_pre.

How does attendance for games relate to elo1_pre?

How accurate are the predictions? We can compare the metrics with their win and lose record?

```{r}

```



```{r}
nfl_elo_MIN <- nfl_elo %>% 
  filter(team1 == "MIN") %>% 
  group_by(season) %>% 
  summarise(total_score = sum(score1)) %>% 
  
   ggplot(aes(y = total_score, x = season))+
  geom_point()+
  geom_line()
```
```{r}
ggplotly(nfl_elo_MIN)
```


Questions for further Research:

We plotted the total scoring of the team per season, and we can look at why the trends occured the way they did.

How does attendence relate to the total scores for each season?

It would also be benefical to exmaine socring differential. 



