---
title: "Demographic Composition the NBA From 1996-2016 Seasons"
author: "Michael Lockhart"
date: "12/17/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, include=FALSE}
library(readr)
library(tidyverse)
library(readxl)
library(maps)
library(ggthemes)
library(ggmap)
library(ggplot2)
library(lubridate)
library(babynames)
library(plotly)
library(gganimate) 
library(gifski) 
```

```{r, message=FALSE, include=FALSE}
all_seasons <- read_csv("all_seasons.csv")
```

```{r, message=FALSE, include=FALSE}
nbaStats <- read_csv("nbaStats.csv")
```

```{r, message=FALSE, include=FALSE}
NBA_Players_by_State <- read_excel("NBA Players by State.xlsx") %>% 
  mutate(State= ifelse(State== "Washingon", "Washington", State))
```

```{r, include=FALSE}
NBA_Players_birthplace <- NBA_Players_by_State %>% 
  select(Player, Date, City, State)
```

```{r, include=FALSE}
dist_players <- all_seasons %>% 
distinct(player_name,.keep_all = TRUE) %>% 
  select(player_name, country, draft_number)
```

```{r, include=FALSE}
pts_state <- all_seasons %>% 
  left_join(NBA_Players_birthplace, by = c( "player_name" = "Player")) %>% 
  mutate(state_new = tolower(ifelse(State=="DC", "district of columbia", State))) %>% 
  group_by(state_new) %>% 
  summarise(avg = mean(pts)) %>% 
  filter(state_new != "NA")
```

```{r, include=FALSE}
totalp <- all_seasons %>% 
  left_join(NBA_Players_birthplace, by = c( "player_name" = "Player")) %>% 
  mutate(state_new = tolower(ifelse(State=="DC", "district of columbia", State))) %>% 
  distinct(player_name,.keep_all = TRUE) %>% 
  group_by(state_new) %>% 
  filter(state_new != "NA") %>% 
  summarise(Total = n())
```

```{r, include=FALSE}
state_data <- tibble(name = str_to_lower(state.name), 
                     code = state.abb) 

total_states <- all_seasons %>% 
  left_join(NBA_Players_birthplace, by = c( "player_name" = "Player")) %>% 
  mutate(state_new = tolower(ifelse(State=="DC", "district of columbia", State))) %>% 
  distinct(player_name,.keep_all = TRUE) %>% 
  group_by(state_new) %>% 
  filter(draft_number != "Undrafted")  %>% 
  mutate(draft_pos = as.numeric(draft_number)) %>% 
  summarise(Avg = mean(draft_pos)) %>% 
  filter(state_new != "NA") %>% 
  left_join(state_data, by= c("state_new" = "name")) %>% 
  mutate(code= ifelse(state_new== "district of columbia", "DC", code)) %>% 
  left_join(totalp, by = "state_new")
```

```{r, include=FALSE}
pts_total_states <- pts_state %>% 
  left_join(total_states, by = "state_new")
```

```{r, include=FALSE}
state_data <- tibble(name = str_to_lower(state.name), 
                     code = state.abb) 

undrft_map <- all_seasons %>% 
  left_join(NBA_Players_birthplace, by = c( "player_name" = "Player")) %>% 
  mutate(state_new = tolower(ifelse(State=="DC", "district of columbia", State))) %>% 
  distinct(player_name,.keep_all = TRUE) %>% 
  group_by(state_new) %>% 
  filter(draft_number == "Undrafted")  %>% 
  summarise(Undrafted = n()) %>% 
  filter(state_new != "NA") %>% 
  left_join(state_data, by= c("state_new" = "name")) %>% 
  mutate(code= ifelse(state_new== "district of columbia", "DC", code)) %>% 
  left_join(totalp, by = "state_new")
```

```{r, include=FALSE}
dist_countries <- all_seasons %>% 
  group_by(season) %>% 
  distinct(country) %>% 
  count(country) %>% 
  summarise(total_countries = sum(n)) 
```

```{r, include=FALSE}
Ttl <- dist_players %>% 
  group_by(country) %>% 
  summarise(Total = n())
```

```{r, include=FALSE}
undrft <- dist_players %>% 
  group_by(country) %>% 
  filter(draft_number == "Undrafted", country != "USA") %>% 
  count(draft_number) %>% 
  left_join(Ttl, by = "country") %>% 
  arrange(desc(Total))
```

```{r, fig.height= 10, fig.width= 8, include=FALSE}
country_mean <- dist_players%>% 
  filter(country != "USA") %>% 
  filter(draft_number != "Undrafted") %>% 
  mutate(draft_pos = as.numeric(draft_number)) %>% 
  group_by(country) %>%  
  summarize(mean = mean(draft_pos), total = n()) %>%
  arrange(desc(mean))
```

```{r, include=FALSE}
pts_country <- all_seasons %>% 
  group_by(country) %>% 
  summarise(avg = mean(pts))

```

```{r, include=FALSE}
total_players <- dist_players %>% 
  group_by(country) %>% 
  summarise(total = n()) %>% 
  filter(country != "USA", country != "U.S. Virgin Islands")

```

```{r, include= FALSE}
total_ptsAVG <- pts_country %>% 
  left_join(total_players, by = "country") %>% 
  filter(country != "U.S. Virgin Islands") %>% 
  arrange(desc(avg))

 
```


# Background 


We decided to center our project around the NBA, also known as the National Basketball Association. The association was founded in June 1946, and is now widely considered to be the premier men's professional basketball league in the world. In the United States and Canada it is one of the four major professional sports leagues, which also includes Major League Baseball, the National Hockey League, and the National Football League. The NBA has a regular season that runs from October to April, during which each of the 30 teams playing 82 games. The NBA players are also some of the world's best paid athletes by average annual salary.

From the beginning of the project, we knew that we wanted to focus on some aspect of the sport world. We realized that in the common vernacular, we hear most about statistics and visualizations when talking about sports. In terms of content, there are hundreds of television and radio programs, magazines, video games, and hundreds of thousands of Internet sites that are primarily focused on sports, Sports are also unique in the fact that they capture the attention of the public on an ongoing basis, and remain in the public sphere. Sports in general are significant to people across the nation and world wide. More generally they provide resources for the construction of "a meaningful social life in relation to a changing societal environment that has the potential to destabilize and threaten these things". Many people in their free time look up sports “stats” because they are emotionally invested in a team or are participating in fantasy leagues. As Robert V. Bellamy Jr., author of the article "Sports Media: A Modern Institution", put it sports have the ability to cross "international and intercultural boundaries despite differences in language and both written and visual literacy" and this is because "game and competition culture essentially is universal.” One does not have to understand all of the rules and strategies to understand the basic point of the game. On an individual level sports have also impacted our lives in a variety of ways. 

From a data science standpoint we wanted to understand the significance of sports in general and how the coding we learned in class could help us understand trends and maybe make predictions. We ended up focusing primarily on the NBA with the majority of our data came from three data sets called “all_seasons”, “nbaStats”, and "NBA Players by State". The "all_seasons" data set contains two decades (1996-2016) of data on each player who has been part of a team's roster. It illustrates demographic variables such as age, height, weight and country of origin. It also capture biographical details like the team they played for, draft year and round. In addition, it also demonstrates statistics like games played, average number of points, rebounds, assists, etc. "nbaStats" looks at player's status, position, label, as well as other various statistics from 1996 to 2016. "NBA Players by State" was mostly useful as it had the states domestic players were born in. It also provided other statistics like the number of years play. We wanted to understand the correlation between players percieved value and their demographics. We used variables like draft position, number of undrafted players, total of players from state/country, and average points scored to capture this percieved value.  


# Foreign Country Representation in the NBA

```{r, include=FALSE}
  total_countries <- ggplot(dist_countries, aes(x = season, y = total_countries))+
  geom_col(aes(fill = "coral1"))+
  theme(legend.position = "none") +
  coord_flip()+
  labs(title = "Number of Countries Represented in the NBA per Season (1996-2016)", x = "NBA Season" , y = "Total Number of Countries") 
```

```{r, echo=FALSE, fig.width=10}
ggplotly(total_countries) 
```

**Figure 1** 
This graph illustrates the total number of countries represented in the NBA. Representation is based on the country of origin of players in the NBA. The dataset is comprised of entries from the 1996-2016 seasons.


In the past 2 decades we have seen a drastic rise in the number of countries represented in the NBA. In 1996, players from 9 countries competed for the NBA championship. That number rose to an all time high of 44 in the 2005 season, an increase of nearly 500% in less than a decade. Representation has since settled in the high 30s and low 40s, finishing out the 2010s with 38 countries in the 2019-2020 season. What we hoped to accomplish with our analysis of the data was an exploration of the different ways this growth has played out for each represented country.

The Figure 1 was derived from the “all_seasons” dataset which spans 20 years of NBA seasons beginning in 1996 and ending in 2016. For this visualization we focused specifically on the country and year variables. The bar graph shows the number of countries represented for every year in the 2 decades included in the dataset. For all datasets concerning foreign countries, we filtered out the US. This is because, although the number of foreign players has risen, they still only make up roughly 75% of players in the NBA. As a result, including the US in our plots would severely skew our data. However, given the prominence of US representation, we later explored the influence of US geography on the performance of players in the NBA.

The growth in foreign representation was most prominent between 1996 and 2005, the aforementioned highpoint for representation. The largest increase between seasons occurred between the 1999 season and the 2000 season. In our search for an explanation for this leap, we found that in 2000 the NBA established a focus on their international outreach and created the program Basketball Without Borders (BWB). Basketball without Borders aims to “unite young basketball players to promote the sport and encourage positive social change in the areas of education, health, and wellness.” The organization brings in NBA players and coaches to train the top youth players from Asia, Europe, Latin America and Africa. Since its foundation this organization has reached more than 3,000 participants in 24 cities in 20 countries on 5 continents. Most impressively, 37 former Basketball Without Borders campers drafted into the NBA. Basketball Without Borders is an example of one of the many stories that contribute to such a significant shift in an international phenomenon such as the NBA. 

# Player Value (Draft Position) 

Following our findings from Figure 1, we explored the correlation of the origin countries of players in the NBA to their assigned value as basketball players. We wanted to look not just at where a player came from, but further into how their origin affected their perceived value. Out  of all the variables present in the all_seasons dataset, draft position did a great job at directly representing the assigned value of a player in the NBA. Our next visualization shows the correlation between birth country and draft position. 

```{r, fig.height= 10, fig.width= 8, include=FALSE}
plot <- country_mean %>%   
  ggplot(aes(x=fct_inorder(country), y=mean)) +
 geom_point(aes(size = total, fill = "coral1"))+
  theme(legend.position = "none") +
  coord_flip()+
  labs(title = "Average Draft Position of Players per Country (1996-2016)",  x= "Country", y= "Average Draft Position")

```

```{r, fig.height=10, fig.width=10, echo=FALSE}
ggplotly(plot)
```




**Figure 2**
This graph shows the average draft position of players selected in the NBA draft based on their country of origin. We used draft position to examine a subjective value of players based on the country of origin. X-axis measures the draft position and the size of the point measures the total number of players that have played in the NBA from that country (1996-2016 seasons). The NBA draft consists of two rounds, each with 30 selection slots for a total of 60 picks. The lower the number of the selection slot, the more valuable the player is viewed.


In order to achieve a comprehensive and understandable representation of the data, this visualization deals with the average draft position for each country. The averages are plotted against their corresponding countries in a point plot, which includes the number of players representing each country as the size of the points on the plot. There are many cases in which a country was only represented by one player in the 2 decades covered by the dataset, so it was important to include the sample size for draft position averages. It quickly became apparent that this was necessary as the US Virgin Islands were the only country with an average draft position of 1. This can be attributed to Tim Duncan, who is often referred to as one of the greatest NBA players of all time and one of the few players to represent the Virgin Islands in the history of the NBA. His career began in 1997 and continued through 2015, nearly exactly the span of the all_seasons dataset. One would expect countries at the extremes of the draft position distribution to have smaller sample size of players, and this is true for the plot.
	
Figure 2 reveals that foreign born NBA players are distributed across all draft positions and follow a roughly standard distribution centered around the expected draft mean. There are no notable outliers, however there is a slight concentration of better represented countries towards the lower end of the distribution. This would suggest there is a slight correlation between better representation and higher draft position averages, but regardless it is a subtle relationship.

```{r, include=FALSE}
total_states
total_states$hover <- with(total_states, paste(str_to_title(state_new), "<br>", "Average Draft Position", Avg ))
  
  
l <- list(color = toRGB("white"), width = 2) 

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white'))
```

```{r, echo=FALSE, fig.width=10}
plot_geo(total_states, locationmode = 'USA-states') %>%
  add_trace(
    z = ~Total, text = ~hover, locations = ~code,
    color = ~Total, colors = 'Blues'
  ) %>%
  colorbar(title = "Total Players in NBA") %>%
  layout(
    title = "Total Number of Players in the NBA per State (1996-2016 Seasons)",
    geo = g
  )
```

**Figure 3** 
This map shows the number of players that have been drafted in the NBA per state in the United States. It is based on a dataset that includes entries from 1996-2016 seasons. The darker the shade, the more players in the NBA from that state. When you hover over a specific state, it shows the average draft position of players from that state. Players that are "Undrafted" are filtered out. 

As previously mentioned, we had filtered out the United States out of our visualizations because the high number of American participants in the NBA distorted the visualized comparison of countries with significantly less players than the United States. Because there is such a high concentration of players within the United States, we thought it would be beneficial to break it down to a state level (see “Number of Players in the NBA per State"). We found that the number of players per state ranges from 0 to 191. California had the most, followed by New York with 110 players. We noticed that the more populated states had the most players, which makes sense. In 2018, California had a total population of around 39.5 million whereas Idaho, which was the birth state of only one player, is home to 1.7 million. Because of its larger population, there is a greater likelihood that state produces more NBA players. Another aspect of the visualization that we noted was that, many of the mountain west and mid-western states had less players overall. Illinois and Michigan seemed to break that trend as mid-western states, having 102 and 73 players respectively. We assumed that this is the case because both Michigan and Illinois have large urban cities with high population density: Chicago and Detroit. 

In this same visualization, we also captured the average draft position per state. As we can see with the “Average Draft Position for Country” the greater the population of the country do not have a strong connection to the average draft position. A similar phenomena is seen within the states of the US. Those states with a high number of players have draft positions in the mid twenties. Again though, with more players you could have very high drafted individuals, along with those that were lower on the list. Most states fall in the 20 position, for this reason. Alternatively, the Dakotas are both home to two players, and they have relatively high average draft positions. South Dakota has the best average, sitting in the fifth position, while North Dakota is higher in the eleventh position. The Bahamas, in terms of countries, is most similar to South Dakota with an average in the 6th position. This data about the Dakotas demonstrates that although there were only two players from the states they were relatively sought after by the league. New Hampshire, on the other hand,  also has a small number of players (only 1) but is in the 45th position. This means that even though this player from New Hampshire was drafted they were seen as less valuable. New Hampshire in this sense is similar to the country of Argentina which has had three players and also averaged around the 45th position. If were to just take this data set into consideration, and were looking for trends, we could assume that there is a better chance of finding a solid player in the Dakotas rather than in New Hampshire. 

## Undrafted Element 

```{r, include=FALSE}
undrft_map
undrft_map$hover <- with(undrft_map, paste(str_to_title(state_new), "<br>", "Total", Total ))
  
  
l <- list(color = toRGB("white"), width = 2) 

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white'))
```


```{r, echo=FALSE, fig.width=10}
plot_geo(undrft_map, locationmode = 'USA-states') %>%
  add_trace(
    z = ~Undrafted, text = ~hover, locations = ~code,
    color = ~Undrafted, colors = 'Oranges'
  ) %>%
  colorbar(title = "Number of Undrafted Players in NBA") %>%
  layout(
    title = "Total Number of Undrafted Players in the NBA per State (1996-2016 Seasons)",
    geo = g
  )
```


**Figure 4** This maps shows the total number of undrafted players in the NBA per state. The map resemebles the total number of players per state Map because we found states with more players in the NBA tend to also have more undrafted players in the NBA. When you hover over a specific state, it also shows the total number of players in the NBA from that state for comparison.


```{r, include=FALSE}
undrft_plot <- undrft %>% 
  ggplot(aes(x = fct_inorder(country), y = n))+
  geom_col(aes(fill= Total))+
  labs(title = "Number of Undrafted Players in the NBA per Country (1996-2016)", x = "Country", y = "Undrafted Players")+
  coord_flip()
  
```

```{r, echo=FALSE, fig.height=10.5, fig.width= 10}
ggplotly(undrft_plot)
```

**Figure 6**
This graph shows the number of players that are "Undrafted" per country represented in the NBA. The NBA draft consists of two rounds, each with 30 selection slots for a total of 60 picks. Undrafted refers to when a player is not slected in the draft but signs with a team as a free agent. "n" refers to the number of players that are undrafted and "Total" refers to the total number of players drafted from a given country.

Although the “Number of Players in the NBA per State" and “Total Number of Undrafted Players” looks similar (with the same states in dark colors), they demonstrate different variables. This differs from the “Number of Undrafted Players in the NBA by Country” visualization where number of players doesn’t seem to have a huge affect on the number of undrafted players. However, the correlation between these two visualizations in the United States makes sense because there is a much higher total number of players per state and you would expect that with more players in general a greater number would be undrafted. Many of the states with few players don’t have any undrafted players like Alaska and Idaho. Those players from the smaller states must have been sought after in their league, and stood out in their state. Interestingly, the player from Hawaii who has a very high point scoring average (see “Average Points Per Game”) was not initially drafted into the NBA but was signed as a free agent. This demonstrates that even if a player is not drafted, it does not mean they are not valuable. 


# Player Value (Points per Game)

Draft position average is a great measurement of a players valued potential, however, potential doesn’t always line up with their performance on court. We decided to further support the correlation between origin country and player performance based on something more tangible. The all_seasons dataset includes a variable that measures the average points per game of every player. To contextualize this within the player’s countries of origin, we took the average of this value across all players from each country. This point plot is constructed with the same representation of sample size as the last figure, which maintains its importance in this visualization as well. The resulting figure shows us The average points per game by country accompanied with the number of players representing each country.

```{r, fig.height=12, fig.width=10, echo=FALSE}
pts_plot <- total_ptsAVG %>% 
  ggplot(aes(x= fct_inorder(country), y= avg)) +
  geom_point(aes(size = total, fill = "coral1"))+
  theme(legend.position = "none") +
  coord_flip()+
  labs(title = "Average Points per Game By Country (1996-2016)", x= "Country", y= "Average Points per Game" )

ggplotly(pts_plot)
```

**Figure 7**
This graph shows the average points per game of players in the NBA based on their country of origin. We used an average of points scored to examine a subjective value of players based on the country of origin. The x-axis measures the average points scored and the size of the point measures the total number of players that have played in the NBA from that country (1996-2016 seasons).

Another way we distinguished the value of players by state was through the average points per game in the two decades, which can be seen in our "Average Points per Game By Country (1996-2016)" visualization. This figure follows a less uniform distribution. There is a concentration around the overall mean, however there is a much longer tail towards the lower end of the distribution. On top of this there are more clear outliers as well. Germany and Senegal are unique in that they are outliers based on the number of players. Senegal had 8 players in the NBA and remains a foreign country with one of the worst point averages. Every other country with a comparably low average had only one player representing them. Germany is on the opposite end of the distribution, with one of the best point averages, yet 6 players representing them in the dataset, a high number for such an extreme.


```{r, include=FALSE}
pts_total_states
pts_total_states$hover <- with(pts_total_states, paste(str_to_title(state_new), "<br>", "Total Players", Total))
  
  
l <- list(color = toRGB("white"), width = 2) 

g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white'))
```


```{r, echo=FALSE, fig.width=10}
plot_geo(pts_total_states, locationmode = 'USA-states') %>%
  add_trace(
    z = ~avg, text = ~hover, locations = ~code,
    color = ~avg, colors = "Greens"
  ) %>%
  colorbar(title = "Points per Game") %>%
  layout(
    title = "Average Points per Game By State (1996-2016 Seasons)",
    geo = g
  )
```

**Figure 8**
This map shows the points per game of players in the NBA based on the state they are from. It is based on a dataset that includes entries from 1996-2016 seasons. The darker the shade, the greater average of points by players from that state.


On the international scale players from most countries on average scored less than 10 points per game, this is simliar to the distribution on a state level in the United States. In general, there are more countries above the 10 point average than there are states. The individual countries, though, have a better chance of getting a higher average with less players. The one player from Hawaii scored on average the same as those from other nations like Italy, Serbia/Montenegro, and Spain who average score 11.9 points per game. One of the lowest averaging states was New Mexico, with an average of only 2.97 points per game, which is similar to the players from the countries of Austria and Senegal. 

Looking primarily at the United States, when compared to the “Number of Players in the NBA per State” visualization, interestingly enough those states with the most players do not score the most on average. The highest scoring state was Hawaii, with an average of about 12 points per game. Meaning that although only one player came from Hawaii, he had an average higher than those players from the rest of the country according to their state. We can also extrapolate that he was probably a good player in general because he was the only one selected to play from his state. 

# Acknowledgments

Because we noticed in our "Number of Countries Represented in the NBA per Season (1996-2016)" visualization that international participation was on the rise, we also wanted to find out if internationalism in the NBA in general was on the rise. We looked at Television viewership to find that out. We found that international viewership was in fact also on the rise, but not in the way we would have expected. While it might be assumed that countries with the most players would have the highest viewership, this was not the case. Kevin Wang, writer for ESPN, noted this as well and stated that “while international fans have a special connection with players from their home countries, that's becoming less important for overall interest.” According to one study in 2014, just 25.5% of total viewership came from the United States with the Philippines, Brazil, Italy and France next on the list. Canada which has its own team, the Toronto Raptors, fell into sixth place overall. In 2017, Everaldo Marques, a sportscaster for ESPN, noticed a similar trend and said that viewership had significantly increased in Brazil compared with the year before, even without Brazilian players like Anderson Varejao and Leandro Barbosa representing the nation. Many believe that although representation of different nations in the NBA is important, the growth in access to technology is the primary factor in the continued growth in viewership. New developments in technology have led to global mass communication, especially the development of satellite television, which has led to unprecedented global audiences.

Even though we found significant data sets there were limitations. For one, the data from "all_seasons" only had the years from 1996 to 2016, so we were limited in our analysis as a whole. The data sets also only included players from the NBA and not from the minor league or the WNBA so we could not capture how those players were valued based on demographics. In doing our analysis we also realized that birth state and birth country are not necesarry the state people are "from", as people regualary move during their life times. In general our topic was made difficult because it was so popular. A lot of data sets that would have been helpful to create visualizations were monetized, and therefore inaccessible to us. We included the data in our written analysis, but were unable to demonstrate the findings in helpful visualization. 

There are many other avenues that could be explore through the data we have collected and the visualizations we created. If we were to further this project, we could look at other ways athletes are valued like how many games they played, how many minutes they played, or even how much money they made in a season. It could also be useful to visualize the information we collected as compared to other major league sports like the NHL, the NFL, or the Women's National Basketball Association. We also would want to look into the factors that might explain why certain geographical areas receive the most players. Number of facilities for players, funding, and sports education would probably be the varibles we would start with. 

Different values of players, rise of new countries, and there were be a rise and value. 


#### Sources
* https://www.nba.com/article/2019/10/22/nba-rosters-feature-108-international-players-2019-20
* http://global.nba.com/basketball-without-borders/
* https://cares.nba.com/international/
* https://flowingdata.com/2014/05/13/nba-fans-by-zip-code/
* https://www.sportsrec.com/8127260/what-is-the-importance-of-sports-in-our-lives
